Full Dissertation

### 2024-04-12 Sections/Introduction.tex
update IGAS intro - then re-render everything

Varied training has been shown to influence learning in wide array of
different tasks and domains, including categorization
\autocite{hahnEffectsCategoryDiversity2005,maddoxStimulusRangeDiscontinuity2011,posnerGenesisAbstractIdeas1968,nosofskyModelguidedSearchOptimal2019,morgensternOneshotCategorizationNovel2019,plebanekEffectsFrequencyVariability2021},
language learning
\autocite{jonesDensityDistinctivenessEarly2020,perryLearnLocallyThink2010,twomeyAllRightNoises2018,wonnacottInputEffectsAcquisition2012,brekelmansDoesHighVariability2022},
anagram completion \autocite{goodeSuperiorityVariableRepeated2008},
trajectory extrapolation
\autocite{fulvioTaskSpecificResponseStrategy2014}, cognitive control
tasks
\autocite{sabahWhenLessMore2019,moshon-cohenStimulusVariabilityImproves2024},
associative learning
\autocite{leeEvidentialDiversityIncreases2019,reichmannVariabilityAbstractionEvaluative2023,fanStimulusDiversityIncreases2022,pradaExperiencedCategoryVariability2020},
visual search
\autocite{georgeStimulusVariabilityTask2021,gonzalezDiversityTrainingEnhances2011,kelleyLearningAttendEffects2009},
voice identity learning \autocite{lavanEffectsHighVariability2019}, face
recognition
\autocite{honigPerceptualSimilarityModulates2022,burtonIdentityVariationRepresentations2016,menonVariationPhotosSame2015},
the perception of social group heterogeneity
\autocite{linvilleExemplarAbstractionModels1993,konovalovaInformationSamplingExplanation2020,parkPerceptionVariabilityCategory1987,gershmanStructureLearningPrinciples2023}
, simple motor learning
\autocite{braunMotorTaskVariation2009,kerrSpecificVariedPractice1978,rollerVariablePracticeLenses2001,willeyLimitedGeneralizationVaried2018},
sports training
\autocite{greenPracticeVariabilityTransfer1995a,northEffectConsistentVaried2019,breslinConstantVariablePractice2012},
and complex skill learning
\autocite{seowTransferEffectsVaried2019,huetEducationAttentionExplanation2011,hacquesVisualControlClimbing2022}.
See \textcite{czyzVariabilityPracticeInformation2021} or
\textcite{ravivHowVariabilityShapes2022} for more detailed reviews.

@@

Further observations consonant with the results of
\textcite{cohenCategoryVariabilityExemplar2001} have since been observed
in numerous investigations
\autocites{hahnEffectsCategoryDiversity2005,perlmanFurtherAttemptsClarify2012,sakamotoPuttingPsychologyBack2008,hsuEffectsGenerativeDiscriminative2010}[but
see][ and
\textcite{yangCategoryVariabilityEffect2014}]{stewartEffectCategoryVariability2002}.
The results of \textcite{sakamotoPuttingPsychologyBack2008} are
noteworthy. They first reproduced the basic finding of participants
being more likely to categorize an unknown middle stimulus into a
training category with higher variability. In a second experiment, they
held the variability between the two training categories constant and
instead manipulated the training sequence, such that the examples of one
category appeared in an ordered fashion, with very small changes from
one example to the other (the stimuli were lines that varied only in
length), whereas examples in the alternate category were shown in a
random order and thus included larger jumps in the stimulus space from
trial to trial. They found that the middle stimulus was more likely to
be categorized into the category that had been learned with a random
sequence, which was attributed to an increased perception of variability
which resulted from the larger trial to trial discrepancies.

>>>>

Varied training has been shown to influence learning in wide array of
different tasks and domains, including categorization
\autocite{hahnEffectsCategoryDiversity2005,maddoxStimulusRangeDiscontinuity2011,posnerGenesisAbstractIdeas1968,nosofskyModelguidedSearchOptimal2019,morgensternOneshotCategorizationNovel2019,plebanekEffectsFrequencyVariability2021},
language learning
\autocite{jonesDensityDistinctivenessEarly2020,perryLearnLocallyThink2010,twomeyAllRightNoises2018,wonnacottInputEffectsAcquisition2012,brekelmansDoesHighVariability2022},
anagram completion \autocite{goodeSuperiorityVariableRepeated2008},
perceptual learning
\autocite{manentiVariabilityTrainingUnlocks2023,lovibondStimulusDiscriminabilityInduction2020,robsonSpecificVariedPractice2022a,zamanPerceptualVariabilityImplications2021},
trajectory extrapolation
\autocite{fulvioTaskSpecificResponseStrategy2014}, cognitive control
tasks
\autocite{sabahWhenLessMore2019,moshon-cohenStimulusVariabilityImproves2024},
associative learning
\autocite{leeEvidentialDiversityIncreases2019,reichmannVariabilityAbstractionEvaluative2023,fanStimulusDiversityIncreases2022,pradaExperiencedCategoryVariability2020,liveseyRevisitingPeakShift2019},
visual search
\autocite{georgeStimulusVariabilityTask2021,gonzalezDiversityTrainingEnhances2011,kelleyLearningAttendEffects2009},
voice identity learning \autocite{lavanEffectsHighVariability2019}, face
recognition
\autocite{honigPerceptualSimilarityModulates2022,burtonIdentityVariationRepresentations2016,menonVariationPhotosSame2015},
the perception of social group heterogeneity
\autocite{linvilleExemplarAbstractionModels1993,konovalovaInformationSamplingExplanation2020,parkPerceptionVariabilityCategory1987,gershmanStructureLearningPrinciples2023}
, simple motor learning
\autocite{braunMotorTaskVariation2009,kerrSpecificVariedPractice1978,rollerVariablePracticeLenses2001,willeyLimitedGeneralizationVaried2018},
sports training
\autocite{greenPracticeVariabilityTransfer1995a,northEffectConsistentVaried2019,breslinConstantVariablePractice2012},
and complex skill learning
\autocite{seowTransferEffectsVaried2019,huetEducationAttentionExplanation2011,hacquesVisualControlClimbing2022}.
See \textcite{czyzVariabilityPracticeInformation2021} or
\textcite{ravivHowVariabilityShapes2022} for more detailed reviews.

@@

Further observations of widened generalization following varied training
have since been observed in numerous investigations
\autocites{hahnEffectsCategoryDiversity2005,perlmanFurtherAttemptsClarify2012,sakamotoPuttingPsychologyBack2008,hsuEffectsGenerativeDiscriminative2010,hoschPriorExperienceVariability2023}[but
see][]{stewartEffectCategoryVariability2002}[ and
\textcite{seitzModelingCategoryVariability2023}]{yangCategoryVariabilityEffect2014}.
The results of \textcite{sakamotoPuttingPsychologyBack2008} are
noteworthy. They first reproduced the basic finding of participants
being more likely to categorize an unknown middle stimulus into a
training category with higher variability. In a second experiment, they
held the variability between the two training categories constant and
instead manipulated the training sequence, such that the examples of one
category appeared in an ordered fashion, with very small changes from
one example to the other (the stimuli were lines that varied only in
length), whereas examples in the alternate category were shown in a
random order and thus included larger jumps in the stimulus space from
trial to trial. They found that the middle stimulus was more likely to
be categorized into the category that had been learned with a random
sequence, which was attributed to an increased perception of variability
which resulted from the larger trial to trial discrepancies.

### 2024-04-12 Sections/full.qmd
update IGAS intro - then re-render everything

Further observations consonant with the results of @cohenCategoryVariabilityExemplar2001  have since been observed in numerous investigations [@hahnEffectsCategoryDiversity2005; @perlmanFurtherAttemptsClarify2012;
@sakamotoPuttingPsychologyBack2008; @hsuEffectsGenerativeDiscriminative2010; but see @stewartEffectCategoryVariability2002 and @yangCategoryVariabilityEffect2014]. The results of @sakamotoPuttingPsychologyBack2008 are noteworthy. They first reproduced the basic finding of participants being more likely to categorize an unknown middle stimulus into a training category with higher variability. In a second experiment, they held the variability between the two training categories constant and instead manipulated the training sequence, such that the examples of one category appeared in an ordered fashion, with very small changes from one example to the other (the stimuli were lines that varied only in length), whereas examples in the alternate category were shown in a random order and thus included larger jumps in the stimulus space from trial to trial. They found that the middle stimulus was more likely to be categorized into the category that had been learned with a random sequence, which was attributed to an increased perception of variability which resulted from the larger trial to trial discrepancies.

@@

Notions of similarity have long played a central role in many prominent models of generalization of learning, as well as in the longstanding theoretical issue of whether learners abstract an aggregate, summary representation, or if they simply store individual instances. Early models of learning often assumed that discrete experiences with some task or category were not stored individually in memory, but instead promoted the formation of a summary representation, often referred to as a prototype or schema, and that exposure to novel examples would then prompt the retrieval of whichever preexisting prototype was most similar [@posnerGenesisAbstractIdeas1968]. Prototype models were later challenged by the success of instance-based or exemplar models -- which were shown to provide an account of generalization as good or better than prototype models, with the advantage of not assuming the explicit construction of an internal prototype [@estesClassificationCognition1994; @hintzmanMINERVASimulationModel1984; @medinContextTheoryClassification1978;
@nosofskyAttentionSimilarityIdentificationcategorization1986 ]. Instance-based models assume that learners encode each experience with a task as a separate instance/exemplar/trace, and that each encoded trace is in turn compared against novel stimuli. As the number of stored instances increases, so does the likelihood that some previously stored instance will be retrieved to aid in the performance of a novel task. Stored instances are retrieved in the context of novel stimuli or tasks if they are sufficiently similar, thus suggesting that the process of computing similarity is of central importance to generalization.

Similarity, defined in this literature as a function of psychological distance between instances or categories, has provided a successful account of generalization across numerous tasks and domains. In an influential study demonstrating an ordinal similarity effect, experimenters employed a numerosity judgment task in which participants quickly report the number of dots flashed on a screen. Performance (in terms of response times to new patterns) on novel dot configurations varied as an inverse function of their similarity to previously trained dot configurations @palmeriExemplarSimilarityDevelopment1997. That is, performance was better on novel configurations moderately similar to trained configurations than to configurations with low-similarity, and also better on low-similarity configurations than to even less similar, unrelated configurations. Instance-based approaches have had some success accounting for performance in certain sub-domains of motor learning [@cohenWhereGraspsAre2004; @crumpEpisodicContributionsSequential2010; @meighWhatMemoryRepresentation2018; @poldrackRelationshipSkillLearning1999; @wifallReachingResponseSelection2017]. @crumpEpisodicContributionsSequential2010 trained participants to type words on an unfamiliar keyboard, while constraining the letters composing the training words to a pre-specified letter set. Following training, typing speed was tested on previously experienced words composed of previously experienced letters; novel words composed of letters from the trained letter set; and novel words composed of letters from an untrained letter set. Consistent with an instance-based account, transfer performance was graded such that participants were fastest at typing the words they had previously trained on, followed by novel words composed of letters they had trained on, and slowest performance for new words composed of untrained letters.

## The effect of training variability on transfer

While similarity-based models account for transfer by the degree of similarity between previous and new experiences, a largely separate body of research has focused on improving transfer by manipulating characteristics of the initial training stage. Such characteristics have included training difficulty, spacing, temporal order, feedback schedules, and the primary focus of the current work -- variability of training examples.

Research on the effects of varied training typically compares participants trained under constant, or minimal variability conditions to those trained from a variety of examples or conditions [@czyzVariabilityPracticeInformation2021; @soderstromLearningPerformanceIntegrative2015]. Varied training has been shown to influence learning in myriad domains including categorization of simple stimuli [@hahnEffectsCategoryDiversity2005; @maddoxStimulusRangeDiscontinuity2011; @posnerGenesisAbstractIdeas1968], complex categorization [@nosofskyModelguidedSearchOptimal2019], language learning [@jonesDensityDistinctivenessEarly2020; @perryLearnLocallyThink2010; @twomeyAllRightNoises2018; @wonnacottInputEffectsAcquisition2012], anagram completion [@goodeSuperiorityVariableRepeated2008], trajectory extrapolation [@fulvioTaskSpecificResponseStrategy2014], task switching [@sabahWhenLessMore2019], associative learning [@leeEvidentialDiversityIncreases2019], visual search [@georgeStimulusVariabilityTask2021; @gonzalezDiversityTrainingEnhances2011; @kelleyLearningAttendEffects2009], voice identity learning [@lavanEffectsHighVariability2019], simple motor learning [@braunMotorTaskVariation2009; @kerrSpecificVariedPractice1978; @rollerVariablePracticeLenses2001; @willeyLimitedGeneralizationVaried2018], sports training  [@greenPracticeVariabilityTransfer1995a; @northEffectConsistentVaried2019], and training on a complex video game [@seowTransferEffectsVaried2019].
Training variation has received a particularly large amount of attention within the domain of visuomotor skill learning. Much of this research has been influenced by the work of @schmidtSchemaTheoryDiscrete1975, who proposed a schema-based account of motor learning as an attempt to address the longstanding problem of how novel movements are produced. According to Schema Theory, learners possess general motor programs for classes of movements (e.g. throwing a ball with an underhand movement), as well as schema rules that determine how a motor program is parameterized or scaled for a particular movement. Schema theory predicts that varied training results in the formation of a more general schema-rule, which can allow for transfer to novel movements within a given movement class. Experiments that test this hypothesis are often designed to compare the transfer performance of a constant-trained group against that of a varied-trained group. Both groups train on the same task, but the varied group practices from multiple levels of a task-relevant dimension that remains invariant for the constant group. For example, investigators might train two groups of participants to throw a projectile at a target, with a constant group that throws from a single location, and a varied group that throws from multiple locations. Both groups are then tested from novel locations. Empirically observed benefits of the varied-trained group are then attributed to the variation they received during training, a finding observed in numerous studies [@catalanoDistantTransferCoincident1984a; @chuaPracticeVariabilityPromotes2019; @goodwinEffectDifferentQuantities1998; @kerrSpecificVariedPractice1978; @wulfEffectTypePractice1991], and the benefits of this variation are typically thought to be mediated by the development of a more general schema for the throwing motion.

Of course, the relationship between training variability and transfer is unlikely to be a simple function wherein increased variation is always beneficial. Numerous studies have found null, or in some cases negative effects of training variation [@deloshExtrapolationSineQua1997; @sinkeviciuteRoleInputVariability2019; @wrisbergVariabilityPracticeHypothesis1987], and many more have suggested that the benefits of variability may depend on additional factors such as prior task experience, the order of training trials, or the type of transfer being measured [@bernikerEffectsTrainingBreadth2014; @braithwaiteEffectsVariationPrior2015; @hahnEffectsCategoryDiversity2005; @lavanEffectsHighVariability2019; @northEffectConsistentVaried2019; @sadakataIndividualAptitudeMandarin2014; @zamanPerceptualVariabilityImplications2021].

@@

Although the benefits of training variation in visuomotor skill learning have been observed many times, null findings have also been repeatedly found, leading some researchers to question the veracity of the variability of practice hypothesis [@newellSchemaTheory19752003; @vanrossumSchmidtSchemaTheory1990]. Critics have also pointed out that investigations of the effects of training variability, of the sort described above, often fail to control for the effect of similarity between training and testing conditions. For training tasks in which participants have numerous degrees of freedom (e.g. projectile throwing tasks where participants control the x and y velocity of the projectile), varied groups are likely to experience a wider range of the task space over the course of their training (e.g. more unique combinations of x and y velocities).
Experimenters may attempt to account for this possibility by ensuring that the training location(s) of the varied and constant groups are an equal distance away from the eventual transfer locations, such that their training throws are, on average, equally similar to throws that would lead to good performance at the transfer locations. However, even this level of experimental control may still be insufficient to rule out
the effect of similarity on transfer. Given that psychological similarity is typically best described as either a Gaussian or exponentially decaying function of psychological distance [@ennisMultidimensionalStochasticTheory1988; @ghahramaniGeneralizationLocalRemappings1996; @loganInstanceTheoryAutomatization1988; @nosofskySimilarityScalingCognitive1992; @shepardUniversalLawGeneralization1987; @thoroughmanRapidReshapingHuman2005 ], it is plausible that a subset of the most similar training instances could have a disproportionate impact on generalization to transfer conditions, even if the average distance between training and transfer conditions is identical between groups. @fig-toy-model1 demonstrates the consequences of a generalization gradient that drops off as a Gaussian function of distance from training, as compared to a linear drop-off.

>>>>

Further observations of widened generalization following varied training have since been observed in numerous investigations [@hahnEffectsCategoryDiversity2005; @perlmanFurtherAttemptsClarify2012;
@sakamotoPuttingPsychologyBack2008; @hsuEffectsGenerativeDiscriminative2010; @hoschPriorExperienceVariability2023; but see @stewartEffectCategoryVariability2002;  @yangCategoryVariabilityEffect2014; and @seitzModelingCategoryVariability2023]. The results of @sakamotoPuttingPsychologyBack2008 are noteworthy. They first reproduced the basic finding of participants being more likely to categorize an unknown middle stimulus into a training category with higher variability. In a second experiment, they held the variability between the two training categories constant and instead manipulated the training sequence, such that the examples of one category appeared in an ordered fashion, with very small changes from one example to the other (the stimuli were lines that varied only in length), whereas examples in the alternate category were shown in a random order and thus included larger jumps in the stimulus space from trial to trial. They found that the middle stimulus was more likely to be categorized into the category that had been learned with a random sequence, which was attributed to an increased perception of variability which resulted from the larger trial to trial discrepancies.

@@

Early models of learning often assumed that discrete experiences with some task or category were not stored individually in memory, but instead promoted the formation of a summary representation, often referred to as a prototype or schema, and that exposure to novel examples would then prompt the retrieval of whichever preexisting prototype was most similar. In addition to being a landmark study on the influence of training variability, @posnerGenesisAbstractIdeas1968 (described above) also put forward an influential argument concerning the nature of the mental representations acquired during learning - namely that learners tend to abstract a prototype, or aggregate representation of the dot pattern categories, rather than encoding each individual stimuli. Recall that participants are trained on only on distortions of the category prototypes (e.g.  low, medium or high distortions), never encountering the exact prototypes during the training stage. Then, in the testing phase, participants are tested with the prototype patterns, their old training items, and novel low, medium and high distortions. The authors found that participants had the highest testing accuracy for the previously unseen prototype patterns, followed by the old training items, and then the novel low, medium and high distortions. The authors interpreted this pattern as evidence that participants had acquired prototype representation of the category, as opposed to storing each individual training instance, and that generalization was based on the similarity of the testing items to the learned prototype representations. @posnerGenesisAbstractIdeas1968 has been extremely influential,  and continues to be cited in contemporary research as clear evidence that prototype abstraction underlies the benefits of varied training. It's also referenced as a key influence in the development of "Schema Theory of Motor Learning" @schmidtSchemaTheoryDiscrete1975, which in turn influenced decades of research on the potential benefits of varied training in motor skill learning. However a number of the core assumptions utilized by @posnerGenesisAbstractIdeas1968 were later called into question both empirically and with competing theoretical accounts [@hintzmanMINERVASimulationModel1984; @hintzmanSchemaAbstractionMultipletrace1986; @palmeriCentralTendenciesExtreme2001; @zakiHighdistortionEnhancementEffect2007; @mcclellandDistributedMemoryRepresentation1985; @nosofskyInvestigationsExemplarBasedConnectionist1992]. @palmeriCentralTendenciesExtreme2001 demonstrated the both the dangers of assuming that psychological representations mimic the metric stimulus space, as well the viability of models with simpler representational assumptions. These authors conducted a near replication of the @posnerGenesisAbstractIdeas1968 study, but also had participants provide similarity judgements of the dot pattern stimuli after completing the training phase. A multidimensional scaling analysis of the similarity judgements revelead that the psychological representations of the prototype stimuli were not located in the middle of the training stimuli, but were instead extreme points in the psychological space. The authors also demonstrated the generalization patterns of @posnerGenesisAbstractIdeas1968 could be accounted for by an exemplar-based model, without any need to assume the abstraction of a prototype.

Instance-based, or exemplar-based models generally assume that learners encode each experience with a task as a separate instance/exemplar/trace, and that each encoded trace is in turn compared against novel stimuli [@estesClassificationCognition1994; @hintzmanMINERVASimulationModel1984; @medinContextTheoryClassification1978; @nosofskyAttentionSimilarityIdentificationcategorization1986; @jamiesonInstanceTheoryDomaingeneral2022]. As the number of stored instances increases, so does the likelihood that some previously stored instance will be retrieved to aid in the performance of a novel task. Stored instances are retrieved in the context of novel stimuli or tasks if they are sufficiently similar, thus suggesting that the process of computing similarity is of central importance to generalization.

Similarity, defined in this literature as a function of psychological distance between instances or categories, has provided a successful account of generalization across numerous tasks and domains. In an influential study demonstrating an ordinal similarity effect, experimenters employed a numerosity judgment task in which participants quickly report the number of dots flashed on a screen. Performance (in terms of response times to new patterns) on novel dot configurations varied as an inverse function of their similarity to previously trained dot configurations @palmeriExemplarSimilarityDevelopment1997. That is, performance was better on novel configurations moderately similar to trained configurations than to configurations with low-similarity, and also better on low-similarity configurations than to even less similar, unrelated configurations. Instance-based similarity approaches have had some success accounting for performance in certain sub-domains of motor learning [@cohenWhereGraspsAre2004; @crumpEpisodicContributionsSequential2010; @meighWhatMemoryRepresentation2018; @poldrackRelationshipSkillLearning1999; @wifallReachingResponseSelection2017]. @crumpEpisodicContributionsSequential2010 trained participants to type words on an unfamiliar keyboard, while constraining the letters composing the training words to a pre-specified letter set. Following training, typing speed was tested on previously experienced words composed of previously experienced letters; novel words composed of letters from the trained letter set; and novel words composed of letters from an untrained letter set. Consistent with an instance-based account, transfer performance was graded such that participants were fastest at typing the words they had previously trained on, followed by novel words composed of letters they had trained on, and slowest performance for new words composed of untrained letters.

@@

Although the benefits of training variation in visuomotor skill learning have been observed many times, null findings have also been repeatedly found, leading some researchers to question the veracity of the variability of practice hypothesis [@newellSchemaTheory19752003; @vanrossumSchmidtSchemaTheory1990]. Critics have also pointed out that investigations of the effects of training variability, of the sort described above, often fail to control for the effect of similarity between training and testing conditions. For training tasks in which participants have numerous degrees of freedom (e.g. projectile throwing tasks where participants control the x and y velocity of the projectile), varied groups are likely to experience a wider range of the task space over the course of their training (e.g. more unique combinations of x and y velocities).
Experimenters may attempt to account for this possibility by ensuring that the training location(s) of the varied and constant groups are an equal distance away from the eventual transfer locations, such that their training throws are, on average, equally similar to throws that would lead to good performance at the transfer locations. However, even this level of experimental control may still be insufficient to rule out the effect of similarity on transfer. Given that psychological similarity is typically best described as either a Gaussian or exponentially decaying function of psychological distance [@ennisMultidimensionalStochasticTheory1988; @ghahramaniGeneralizationLocalRemappings1996; @loganInstanceTheoryAutomatization1988; @nosofskySimilarityScalingCognitive1992; @shepardUniversalLawGeneralization1987; @thoroughmanRapidReshapingHuman2005], it is plausible that a subset of the most similar training instances could have a disproportionate impact on generalization to transfer conditions, even if the average distance between training and transfer conditions is identical between groups. @fig-toy-model1 demonstrates the consequences of a generalization gradient that drops off as a Gaussian function of distance from training, as compared to a linear drop-off.
